Biden Projected to Win the 2024 US Presidential Elections

### 2024-03-16 paper/paper.qmd
other
Cleaning instructions added in appendix

# Introduction
A key turning point in the political environment of the United States occurred with the 2020 presidential election,
which was marked by extreme conflict and serious economic and social challenges. We analyze data from the American
Community Survey (ACS) 2022 sample, which offers a thorough understanding of demographic, socioeconomic,
and geographic aspects across the nation, in order to comprehend the patterns underlying voting choices during
this 2024 election period. This paper focuses on the two major political parties, Democratic and Republican.

The rivalry between the Democratic and Republican parties, as embodied by their opposing candidates,
Joe Biden and Donald Trump, has been an iconic feature of the modern American political scene. The Democratic Party's
nominee is the veteran politician and former Obama vice president Joe Biden. With an emphasis on middle-class families,
Biden's program prioritizes measures that advance racial justice, healthcare reform, climate action, and economic
regeneration. He is an advocate of a more compassionate and inclusive style of governing, hoping to bring the
country back together and bring decorum back into political conversation.

On the other hand, the current Republican Party standard-bearer and President Donald Trump represents a more populist
and nationalist agenda. A major focus on economic nationalism, immigration restrictions, tax cuts, and
deregulation have been highlights of Trump's leadership. His administration prioritized themes like law and order,
trade protectionism, and border security as part of its conservative agenda. Trump's aggressive approach to politics
is a defining characteristic of his leadership style; it frequently divides public opinion and provokes strong emotions
from both supporters and opponents.

Our estimand is the relationship between demographic, socio-economic, and geographic variables (such as age, income,
education, state of residence, and gender) and voting behavior during the 2022 presidential election.
Using this extensive dataset, our goal is to identify trends and factors that influence voting behaviour,
clarifying the complex relationship between personal traits and local factors that influences election results.

Our research adds to a greater understanding of the factors influencing voting decisions through comprehensive
statistical analysis, yielding insightful data that may influence future election campaigns and policy discussions.
Through highlighting the subtleties of voting patterns in the 2022 election, our research provides a critical insight
on the condition of American democracy and the factors motivating involvement in politics and engagement.

The paper is further organised into three sections: Data, Model, Results, and Discussion. In the Data section, we
discuss how the datasets to be used for the analysis were obtained and pre-processed. We will explain the variables of
interest in the datasets for the analysis. The Model section describes the model being used for the analysis. The Results
section will then highlight and discuss the trends and associations found during the analysis. Lastly, the Discussion
section will talk about some interesting trends found in Results in depth, link it to the real world and also highlight
the weaknesses of our analysis.

@@

```
In @fig-support-state we see the percentage of the votes for Biden/Democratic party. This graph uses the data we
collected after the post-stratification. Here, the red dots on the graph represent the voting percentage and the red lines are the error bars. The black vertical line with dashes indicates the line of proportion. Here, it is the equal split between support for Biden and support for the opposing candidate, Trump.

@@

# Discussion

## Insights Gained:

Our analysis, based on sample data from the 2022 survey, followed by post-stratification focusing on key variables, Age,
State, and Sex, yielded valuable insights into voter behavior. Notably, our findings indicated an inclination within the
population towards voting for Biden from the Democratic Party.

In the graph showcasing different regions and their voting for Biden, we noticed that
this is similar to the findings in an article by statistician James E. Campbell,  [@James_2016] that address the same prediction for 2024. Our model and the
article predict that states that support Trump are from the middle and Midwest parts of the US, whereas coastal states
and the Northeast tend to vote for Biden.

Historically, rural, conservative-leaning towns with strong cultural ties to traditional values and industries like
manufacturing and agriculture have defined the middle and Midwest regions of the United States.
These areas frequently support gun rights, minimal government intrusion, and religious liberty—issues that are strongly
aligned with the agenda of the Republican Party, which Trump advocated. In contrast, the Northeast and coastline states,
which are renowned for their progressive views on culture, urban facilities, and diverse populations, have historically
tended toward more liberal policies and social ideals that are more in line with the Democratic Party platform, which Biden advocated.

We also created a graph highlighting different age groups and what their voting opinions are. We noticed that
the youngest group of voters (age 18-24). The emphasis on social justice and LGBTQ+ rights among the younger generation
may be the cause of this difference[@Brownstein_2023]. Since there are undoubtedly more young people who identify as sexual minorities,
there is a trend for them to support Biden and the Democratic Party. We also noticed that the older generation i.e
individuals of age group higher than 60 years old were inclined to voting for Biden as well.
National security and conservative principles, which are frequently connected to the Republican Party,
may have an impact on the elder generation.

Lastly, our third variable of study was Sex. We noticed a slight bias towards the democratic party from the females than
the male gender. Women are predicted to vote for the democratic party 0.6 times more than men.
Studies have consistently demonstrated that there is a gender difference in political ideology, with women generally
favoring liberal or progressive policies over those of males.[@Igielnik_2021]
This discrepancy is sometimes linked to the fact that women prioritize issues more than men do, with women giving social
welfare, healthcare, education, and reproductive rights—all of which are more in line with the Democratic Party program.
With its focus on women's rights, healthcare reform, and inclusivity, Biden's campaign may have connected more
deeply with female voters, resulting in higher levels of support.

## Comparative Analysis:
Our findings align with existing literature on electoral behavior, which underscores the significance of demographic
variables in shaping voter preferences. While our study offers insights into the inclination towards voting for Biden,
it is important to contextualize our findings within the broader landscape of political science research.[Gelman]

Our analysis contributes to a growing body of evidence highlighting the complex interplay between socio-economic factors
and political attitudes, providing valuable insights for future research and decision-making in the political arena.

## Contributions to Knowledge:
By focusing our analysis on specific demographic variables, we contributed to a deeper understanding of the underlying
determinants shaping electoral outcomes. Our study corroborates existing literature suggesting that demographic factors
such as age, income, education, and geographic location play pivotal roles in influencing voter preferences. [@Gelman]
Moreover, our utilization of post-stratification techniques allowed for a more granular examination of voting behavior,
offering insights into the heterogeneity of voter sentiment across different demographic groups.

## Weaknesses and Limitations:

While our analysis provides valuable insights into voter behavior, it is important to acknowledge several limitations. 
Firstly, our reliance on sample data from the 2022 survey may introduce biases and inaccuracies in our predictions.
Our reliance on sample data from the 2022 survey may introduce a sampling bias, as the sample might not fully represent
the diversity of the electorate. Consequently, this could lead to inaccuracies in our predictions and limit the
generalizability of our findings to the broader population.

Additionally, while we focused on key demographic variables, other influential factors such as political ideology and
campaign messaging were not fully accounted for in our model. This means that our research has the inaccuracy of
Incomplete Variable Consideration[@Funakoshi_2020].

Furthermore, uncertainties in the post-stratification process may impact the reliability of our estimates,
warranting caution in interpreting the observed inclination towards voting for Biden.

## Future Directions:
Moving forward, further research is warranted to address the limitations identified in our study and advance our
understanding of electoral dynamics. Future studies could explore the integration of additional data sources,
such as social media sentiment analysis and polling data, to enhance the accuracy of election predictions.
Additionally, efforts to improve transparency and reproducibility in data collection and analysis are essential
for enhancing the reliability and validity of electoral forecasts.

@@

## Cleaning

>>>>

# Introduction

@@

In @fig-support-state we see the percentage of the votes for Biden/Democratic party. This graph uses the data we
collected after the post-stratification. Here, the red dots on the graph represent the voting percentage and the red lines are the error bars. The black vertical line with dashes indicates the line of proportion. Here, it is the equal split between support for Biden and support for the opposing candidate, Trump.

@@

@@

## Cleaning
For the survey dataset, the cleaning steps we took were:

1. Initial cleaning involved filtering the dataset to retain only responses relevant to the analysis. Responses were filtered based on the participant's stance on Q1_1 which is “Overall, do you have a favorable or unfavorable impression of...Donald Trump?”, specifically selecting those labeled as "Favorable" or "Unfavorable".

2. Age groups were defined based on the participants' reported age (ppage). Participants were categorized into four age groups: "18-29", "30-44", "45-59", and "60+". The final was named age.

3. Respondents' income information (ppinc7) was retained in its original format to provide insights into income distribution. The final column was named income.

4. Education levels were standardized to facilitate analysis. Respondents' education information (ppeducat) was categorized into three groups: "High school", "More than high school" (including "Bachelors degree or higher" and "Some college"), and "Less than high school". This was done to match the education level separation available in the poststratification dataset. The final column was named education.

5. The state of residence for each participant (ppstaten) was preserved for regional analysis. The final column was named state.

6. Gender information (ppgender) was just had a change of turning the value to lower case. The final column was named sex as just having these two criterias implies that and to maintain with the poststratification data reporting sex as well.

7. We created a binary variable for favorability of Biden. Responses from Q1_1 labeled as "Favorable" were coded as 0, while responses labeled as "Unfavorable" were coded as 1. The final column was named vote_biden.

8. The final survey dataset contained the above mentioned 6 columns: age, education, income, state, sex, and the coded variable for Biden vote.

For the poststratification dataset, the cleaning steps we took were:

1. Initially, the "AGE" variable was converted to numeric format. Subsequently, data points were filtered to retain only those individuals aged 18 and above, excluding individuals from Vermont and Montana and cases where total household income (FTOTINC) was recorded as less than zero. Individuals from Vermont and Montana were excluded from the dataset to maintain consistency with the survey data as the survey data does not involve any records from these two states.

2. Participants' ages (AGE) were categorized into four groups: "18-29", "30-44", "45-59", and "60+". The final column was named age.

3. Education levels were standardized into four categories: "High school", "More than high school", and "Less than high school". This categorization was based on the reported education level (EDUC), with corresponding numerical codes for each category. The final column was named education.

4. States were labeled according to their corresponding numerical codes (STATEICP), assigning each code to its respective state name. The final column was named education.

5. Sex information (SEX) was changed for the codes to their respective values: "male" and "female". The final column was named sex.

6. Household income (FTOTINC) was categorized into the same division of income brackets as the survey data: Less than $10,000, $10,000 to $24,999, $25,000 to $49,999, $50,000 to $74,999, $75,000 to $99,999, $100,000 to $149,999 and $150,000 or more. The final column was named income.

7. The final poststratification dataset contained the above mentioned 5 columns: age, education, income, state and sex.

Now we are ensured that both the datasets have the same variable names and respective categories.
